Generalizing the Model View Controller Paradigm

### 2006-07-29 inst/paper/MVC.tex
Made changes to paper 2 and also updated the vignette.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@18994 bc3139a8-67e5-0310-9ffc-ced21a209358

\usepackage{epsfig}
\usepackage{comment}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{times}
\usepackage[authoryear,round]{natbib}

@@

There are many ways to visualize multivariate data when more than
two or three dimensions are needed to view the data. Many authors
(citations) have suggested that linked, interactive views of the data
provide a valuable and important set of tools for comprehending such
data. The model view controller (MVC) paradigm is now a standard
mechanism for implementing such systems. The paradigm has proven very
effective and allows for clear separation of different components.
In this paper we consider an extension of the MVC paradigm to allow
for, what we believe to be, an easy expansion to linked data sets.

The sorts of problems we are interested in, are those where a user
might begin with a large data set, identify some interesting subset
and want to create interactive views of that subset. In some cases the
natural data representation of the subset is in a form other than that
of the initial data set. This leads to complications in the usual MVC
paradigm, where the data (i.e. the model) are often presumed to be in
a standard format. We found that adding new views under such
conditions could be quite problematic and often leads to very complex
code. Hence we considered an approach that created new instances of
the MVC that were linked to the parent that spawned them.

In particular, our implementation consists of a hierarchy of MVCs,
stored as a tree. Each parent can have multiple children, but the
children can have a single parent. Communication between parent and
child is bidirectional, and a variety of different filters can be used
to control flow. Thus, we can devise a number of purpose built MVC
classes, each with a model appropriate to the specific data of
interest, that are linked in a natural way. Interacting with the view
in a child MVC, first sends a message to the child MVC, which will
update all of its views, but the message will also propagate to the
parent MVC (and further up to the root). The parent MVC can choose to
react, or not, to messages from its children -- all control being in
the hands of the users, should they want it.

@@

%% FIXME: perhaps we could have a heatmap and a few of the other views
%% here?
%% EAW: put them below
There is often substantial interest in relating gene expression data
to such a graph based model, or to chromosomal locations, or to
activity in known biological pathways. Data mappings for this latter
aspect can be obtained from different sources such as \cite{KEGG}.
Further, users typically want to begin with a heatmap view. Then from
that derive a GO graph view, and additionally a chromosomal location
view. In order to interact with the GO graph, the primary unit is the
node in that graph (and sometimes one expects with the edges), while
for the heatmap view the basic element is generally the gene.  In Figures
\ref{Fig:heatmap} and \ref{Fig:GOgraph}, a heatmap view of gene expression
from Chronic Lymphocytic Leukemia (CLL) patients and the corresponding GO
graph are shown, respectively.  By interacting with the heatmap to color one
gene blue (shown as gene PDE8A on the heatmap), which causes the corresponding
GO nodes to be colored blue, one can easily see how this gene relates to the
GO terms.

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=4.3in, width=4.3in]{GOgraph.jpg}
    \caption{ Gene Ontology (GO) graph that is associated with the genes
      displayed in the heatmap.  Notice that one gene (PDE8A) is colored blue
      in the heatmap and all of the corresponding GO terms are colored blue in
      the graph. }
    \label{Fig:GOgraph}
  \end{center}
\end{figure}

@@

By \textit{linked data sets} we mean that the data have more than one
conceptual grouping and that there is a relationship between these
different conceptual groupings.  Linked data sets are 
familiar to users of relational databases.  Data in a relational
database are typically stored in many linked tables with rows in
different tables linked to each other through a unique identifier
called a key.  In this example, each table can be considered a data set
and they are linked through the keys.  But the idea is more general
and we shall discuss other situations where the linked data sets do
not easily correspond to tables, however, we must have a map that goes
from the entities in one data set to those in the other as well as the
inverse map, in order to propagate signals between the MVCs.

Consider Figure \ref{Fig:firstMP}, which provides a cartoon
description of the steps that occur when a user interacts with a
view. In the cartoon there are only two views, one of data set A and the
other of data set B, which is linked to data set A.  When a user
interacts with the scatter plot view of data set B, the response to
this interaction is for data set B to update itself (this is indicated
on the figure as a circled 1).  In Step 2, the scatter plot view is
told to update itself.  Next, in Step 3, because data set B is linked
to data set A, data set B sends a message to data set A.  In Step 4,
the information from data set B is converted into information that
data set A understands.  In Step 5, this information is received by
data set A and data set A updates itself.  Finally, in Step 6, data
set A tells its view to update itself.

@@

%%FIXME: please update the figure so that it has A and B, not 1 and 2
%%for the data sets - it was too confusing
%%EAW: it's changed
\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=3.3in, width=5in]{firstMessagePassing.jpg}
    \caption{ Linked views of linked data sets.  When a user interacts with
      the y versus x scatter plot, the first step, indicated by a circled one,
      is to tell data set B to update.  Then all views of data set B must be
      updated.  Next any data that are linked to data set B must be notified
      that data set B changed.  This information from data set B must be
      altered so that it can be understood by linked data sets, which is shown
      in the fourth step.  Finally, data set A is notified that it should
      update in response to the change in data set B and then all views of
      data set A are updated. }
    \label{Fig:firstMP}
  \end{center}
\end{figure}

@@

In essence the MVC design assumes that there is one data set, which is
called the model, that all views reference.  But problems arise when
the data consist of more than one conceptual grouping.  For example,
consider data stored in a relational database, where there are
multiple tables of data and rows in different tables are linked
through a unique identifier called a key.  Although these tables of
data could be combined into one large table, there are several reasons
not to do this.  Keeping the data in separate tables reduces
redundancy, particularly when the rows are related by a one-to-many or
a many-to-many relationship.

@@

Consider Figure \ref{Fig:DBTab} shown below.  Here the database
consists of three tables, State, City and StateCity.  The State table
has three columns, State (the key), Life Expectancy, and Income.  The
City table has three columns, City (the key), Precipitation, and
Temperature.  The StateCity table has two columns, State and City.
This last table relates which cities are in which states.  In this
example, each city can be in only one state, but each state can have
more than one city so there is a one-to-many relationship between
states and cities.  In order to support interactive views of all data
tables, in a standard MVC paradigm, it seems that either all data must
be combined into a single large table (Figure~\ref{Fig:OneDBTab}), or
the processing and accessing of the model data must become much more
complex.

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=0.9in, width=3.5in]{newDatabaseTables.jpg}
    \caption{ Example of Relational Database Tables. }
    \label{Fig:DBTab}
  \end{center}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=1.2in, width=1.5in]{newOneDatabaseTable.jpg}
    \caption{ Example of a Combined Table. }
    \label{Fig:OneDBTab}
  \end{center}
\end{figure}

We felt that neither approach had much appeal. A single large table
suggests that all data are similar, and still leaves issues about how
to deal with ensuring that when, for example, a state is selected all
cities within it are highlit. The second approach requires writing a
large amount of code, but is likely to provide little
extensibility. We should also mention that our main area of
application is a setting where one representation of the data is as a
graph (with nodes and edges) while the other is as a matrix. We can
envisage lots of different MVCs that operate on graph based data, and
equally MVCs that operate on matrix based data. Specifying a mechanism
by which these can interact allows us to create two visible classes
that others can use, but still provide interaction between
appropriately linked instances of the two classes. We also argue that
there are surely other specific types of models that represent
specific types of data (e.g. time series) and it would make sense to
implement these as yet another separate, specialized MVC, that could
be linked to any one of a variety of other MVCs.

@@

As discussed in Section \ref{Sec:Intro}, the MVC design typically uses a single
conceptual unit, such as a data frame or a matrix, for the model.  And
we believe that such an approach is a good one. MVCs should be
specialized to certain types of data, as this allows them to provide
better and more natural interactions. However, we additionally argue
that there is great benefit to considering linking these different
MVCs when the underlying data themselves are derived from the same source.

Each \Robject{MVC} is a coherent object and the interactions between
\Robject{MVCs} are encapsulated in a set of messages.  We use one
\Robject{MVC} object to handle each data set.  Thus, within a
\Robject{MVC} object, one can create linked views of that particular
data set, using the standard MVC design.  Between \Robject{MVC}
objects we propose a second messaging system that allows messages to
be sent between linked data sets.

@@

In Figure \ref{Fig:MultMVC} we show one view for state-level data and
one view for the city-level data.  The state and city data are linked
based on the color.  Even though there are two \Robject{MVC} objects
here, one \Robject{MVC} for each data set, the views are linked.

@@

We believe that there will often be a hierarchical relationship
between the data sets and the MVCs that contain them. We have found
that allowing the user to create new MVCs based on a current MVC is an
important and useful paradigm.  The original MVC is the parent
MVC and the new MVC the child.

@@

In the MVC design, the model stores the data and among other things it
supports all updating and other mechanisms that might be needed.
Information can be sent to the model to add, update, and remove data
and after every change the model is responsible for sending messages
to the views indicating which changes have occurred.  In our
implementation we created a single virtual class that represents an
abstract model and that has storage locations for data that will be
needed by all models. Then, we created a number of subclasses, one for
each type of model. In Figure~\ref{Fig:Model} we show the inheritance
structure for the three types of models we have implemented, one based
on data frames, one based on graphs and one based on the expression
set class of data used by the Bioconductor project \citep{BIOC}.
%%FIXME: we need to send the complete paper2 to the BioC tech report
%%series and then cite it here, saying something like: Complete
%%implementation details are provided in XXX.

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=3in, width=4.7in]{ModelClass.jpg}
    \caption{ Inheritance for Model Objects. }
    \label{Fig:Model}
  \end{center}
\end{figure}

@@

The view classes represent the visual depictions, or renderings, of
the model.  There are some data that are required by all views and all
must respond to a certain subset of events so we have implemented the view
classes by first creating a single, virtual class which the specialized
views extend.  As with our model classes, this allows us to define
methods and data storage once, for the virtual class, and only extend
it for views that specifically require additional data storage, or
specialized handling.  The object model for the view classes is shown
in Figure \ref{Fig:View}.

@@

The controller performs two tasks: the first task is to store
information and the second task is to allow the user to determine the
course of actions.  The information that must be stored primarily
pertains to state. In order for the user to be able to alter the
effects of different interactions, such as changing the response to a
key-click event, or the appearance of the views, the controller must
retain information as to the state.

For the second task, the controller gives the user restricted access
to the state information through an application program interface
(API). In our implementation users can interact either through a GUI
or through command line functions.  Through the API the user can load
data, create views, set the response to user interaction with a view,
quit the program and perform other functions.  On the GUI, there are
menus to let users decide which action is taken next.  If the user
would rather not use the GUI, there are command line functions that
provide access to the same functionality that is available through the
GUI.

@@

Because we support simultaneous linked \Robject{MVC}
objects, more than one \Robject{MVC} object can be loaded and hence
more than one model can be present. However, the control window menus
and the command line functions only perform operations on one
\Robject{MVC} object at a time.  Thus, we make use of the concept of
an \textit{active} \Robject{MVC} object, where the active
\Robject{MVC} object (and thus, the active model) means that
operations performed through the control window menus or through the
command line functions affect that \Robject{MVC} object. Users can
switch between active models, but since the interactions could be
quite different, there must always be an active MVC.

In our implementation we provide a control window, for users that want
to interact via the GUI.  It has menu items corresponding to
\texttt{File}, \texttt{Edit}, \texttt{Display}, \texttt{Event}, and
\texttt{Window}.  These menus always appear on the control window, but
the items under each menu may change depending on the type of model
that is active.  For example, the \texttt{Display} menu allows users
to create views of the active model, but only certain views are
appropriate for a type of model, as mentioned in
Section~\ref{Ssec:OneViews}.  If the active model was a data frame
model (\Rclass{dfModel}), then the views that the user could create
are a scatter plot (\Rclass{sPlotView}) and a spreadsheet
(\Rclass{spreadView}), but a graph plot (\Rclass{graphView}) does not
make sense.  Thus, when the model in the ``active'' \Robject{MVC} is
of class \Rclass{dfModel}, then the Display menu has menu items to
create a scatter plot and a spreadsheet.  Also, if the ``active''
\Robject{MVC} is of class \Rclass{dfModel}, then the command line
functions only allow the user to create a scatter plot or a spreadsheet
view of the active model, since these are the only ones we have
implemented.

Because one of our goals is to create a software package that is
flexible and extensible, users are able to add menus and menu items to
the control window through the command line to access new
functionality that they have created.  As an example if a future
programmer wants to add a new type of view, such as a histogram, then
a new menu item for that view could be added to the \texttt{Display}
menu.  By allowing users to change the controller, they can add new
functionality that is specific to their problem.  Note that when users
make additions to the API, they should add new command line functions
in addition to adding menus and menu items to access that
functionality.

@@

Linking callback functions to events, such as a button click, a mouse
movement or a key press event, is what allows users to have an
interactive environment.  When an event occurs, the callback function
linked to the event determines the response to that event.  To create
a flexible environment it is important to be able to change the
response, to an event.  For example, when a left button press event
occurs, a user may sometimes want a point to be colored and at other
times she may want a point to be hidden.  However, different responses
to an event may require functions with different signatures.  In our
example coloring a point requires two parameters, the new color to use
and the point to color, while hiding a point
requires just one parameter, the point to hide.  Since the intention is
to be able to interchange the responses to an event, the signatures of
different callback functions should be the same.

@@

%%FIXME: I have tried to reword this part, but am not sure I have it
%%right, nor that I am following the design decision here.
%% EAW: When we decided that we wanted callback functions to be
%% interchangeable, we decided that the callback functions must have the same
%% signature so that calls to different callback functions would be
%% identical, as we wouldn't know ahead of time what function we would be
%% calling.  All callback functions must know which object in the plot to
%% act on, but there was no other universal information that all callback
%% functions needed.  Thus, for any other information that a callback function
%% would need, preprocessing functions were used.  As for that one parameter
%% that the callback function needed, the parameter type may be different
%% depending on the plot (and model) type - whatever is considered the object
%% of interest in a plot (and model).  Thus, for plots of data frames,
%% the object in the plot is considered a row from the data frame and for
%% plots of graphs, the object in the plot is considered a node in the graph. 
In our design a callback function takes only one parameter. Further,
within a specific MVC all callback functions must have the same
signature so that different callback functions can be interchanged to
allow different responses to an event.  The class of that parameter
depends on the type of model represented by the \Robject{MVC}.  For
example, for a graph model, the parameter corresponds to a node in the
graph, while for a data frame model, the parameter is of class
\Rclass{character}, representing a case.

%%FIXME: I thought we talked about this, and found it to be a little
%%problematic, as the way this sounds every click will ask me to set a
%%color. Somehow, some of these are essentially global options that
%%need to be set once, and then used.
%% EAW: The preprocessing functions are called once when the callback function
%% is linked to the event - they are not called each time the event occurs.
Any other information a callback function may need is set using the
preprocessing functions, which are called once when the callback function is
linked to the event.  The preprocessing functions take no
parameters.  The purpose of the preprocessing functions is to set
variables that the callback function needs.  Thus, not all callback
functions need preprocessing functions and if no preprocessing
functions are needed, then this slot is NULL.  If the callback
function sets the color of a node in a graph, then a preprocessing
function is needed to determine what color is used.  So in this
example, the preprocessing function opens a color browser to allow the
user to choose the new color.

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=3.4in, width=4.5in]{newMessageClass2.jpg}
    \caption{ Inheritance for Message Classes that are Passed Between
      \Robject{MVCs}. }
    \label{Fig:BetwMess}
  \end{center}
\end{figure}

>>>>

\usepackage{epsfig}
\usepackage{comment}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{times}
\usepackage[authoryear,round]{natbib}
\usepackage{subfigure}

@@

There are many ways to visualize multivariate data when more than
two or three dimensions are needed to view the data.  Many authors
\cite{intGrUnwin} \cite{iPlots} \cite{MONDRIAN} \cite{GGobi}
have suggested that linked, interactive views of the data
provide a valuable and important set of tools for comprehending such
data.  The model view controller (MVC) paradigm is now a standard
mechanism for implementing such systems.  The paradigm has proven very
effective and allows for clear separation of different components.
In this paper we consider an extension of the MVC paradigm to allow
for, what we believe to be, an easy expansion to linked data sets.

The sorts of problems we are interested in, are those where a user
might begin with a large data set, identify some interesting subset
and want to create interactive views of that subset.  In some cases the
natural data representation of the subset is in a form other than that
of the initial data set.  This leads to complications in the usual MVC
paradigm, where the data (i.e. the model) are often presumed to be in
a single, standard format.  We found that adding new views under such
conditions could be quite problematic and often led to very complex
code.  As we shall demonstrate, there are also issues that arise due to
mappings that are not one-to-one between elements on different views.
Hence we considered an approach that created new instances of
the MVC that were linked to the parent that spawned them.

In particular, our implementation consists of a hierarchy of MVCs,
stored as a tree.  Each parent can have multiple children, but the
children have a single parent.  Communication between parent and
child is bidirectional, and a variety of different filters can be used
to control flow.  Thus, we can devise a number of purpose built MVC
classes, each with a model appropriate to the specific data of
interest, that can be linked in a natural way.  Interacting with the view
in a child MVC, first sends a message to the child MVC, which will
update all of its views, but the message will also propagate to the
parent MVC (and further up to the root). The parent MVC can choose to
react, or not, to messages from its children -- all control being in
the hands of the users, should they want it.

@@

%% FIXME: perhaps we could have a heatmap and a few of the other views
%% here?
%% EAW: put them below
There is often substantial interest in relating gene expression data
to such a graph based model, or to chromosomal locations, or to
activity in known biological pathways. Data mappings for this latter
aspect can be obtained from different sources such as \cite{KEGG}.
Further, users typically want to begin with a heatmap view. Then from
that derive a GO graph view, and additionally a chromosomal location
view. In order to interact with the GO graph, the primary unit is the
node in that graph (and sometimes one expects with the edges), while
for the heatmap view the basic element is generally the gene.  In Figures
\ref{Fig:heatmap} and \ref{Fig:GOgraph}, a heatmap view of gene expression
from Chronic Lymphocytic Leukemia (CLL) patients and the corresponding GO
graph are shown, respectively.  By interacting with the heatmap to color one
gene blue (shown as gene PDE8A on the heatmap), which causes the corresponding
GO nodes to be colored blue, one can easily see how this gene relates to the
GO terms.  Also by interacting with the node with an arrow next to it in
Figure \ref{Fig:GOgraph}, which represents the signal transducer activity GO
term, one can create a heatmap view that shows only the genes annotated at
that node, as shown in Figure \ref{Fig:heatmapsub}.  Notice how all genes
annotated at the signal transducer activity GO term have a blue label in
Figure \ref{Fig:heatmapsub}, which shows that this view was derived from a
blue colored node in Figure \ref{Fig:GOgraph}.  Compare the heatmaps in
Figures \ref{Fig:heatmapsub} and \ref{Fig:heatmap}.  The heatmap in Figure
\ref{Fig:heatmapsub} is just a subset of the heatmap in Figure
\ref{Fig:heatmap}, however in Figure \ref{Fig:heatmapsub} eleven genes have a
blue label and in Figure \ref{Fig:heatmap} only one gene has a blue label.
This difference in these heatmaps indicates that views are synchronized based
on the parent-child relationships.  Figure \ref{Fig:heatmap} is the parent of
Figure \ref{Fig:GOgraph}, which is the parent of Figure \ref{Fig:heatmapsub}.
Figures \ref{Fig:heatmap} and \ref{Fig:GOgraph} are synchronized, and Figures
\ref{Fig:GOgraph} and \ref{Fig:heatmapsub} are synchronized, but Figures
\ref{Fig:heatmap} and \ref{Fig:heatmapsub} are not synchronized by color
because they do not have a parent-child relationship: Figure \ref{Fig:heatmap}
is the grandparent of Figure \ref{Fig:heatmapsub}.

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=4.3in, width=4.3in]{GOgraph2.jpg}
    \caption{ Gene Ontology (GO) graph that is associated with the genes
      displayed in the heatmap in Figure \ref{Fig:heatmap}.  Notice that one
      gene (PDE8A) is colored blue in the heatmap and all of the corresponding
      GO terms are colored blue in the graph.  The arrow next to one of the
      nodes indicates that this node was used to create the view shown in
      Figure \ref{Fig:heatmapsub}.  The node with the arrow next to it
      represents the signal transducer activity GO term. }
    \label{Fig:GOgraph}
  \end{center}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=5in, width=5in]{heatmapsub.jpg}
    \caption{ A heatmap view that shows only the genes that are annotated at
      the node in Figure \ref{Fig:GOgraph} that has an arrow next to it.  This
      node represents the GO term, signal transducer activity, and as shown in
      this heatmap, eleven genes are annotated at this node.  Notice that all
      the labels in this heatmap are colored blue, indicating that the parent
      node that was used to create this view was also colored blue.  
      % should I also mention that this is different from the genes that are
      % selected in the grandparent model, which is shown in Figure
      % \ref{Fig:heatmap}??  These eleven genes are also shown in the
      % grandparent view, but only one of them is colored blue showing that
      % views are synchronized by parent-child relationships.
     }
    \label{Fig:heatmapsub}
  \end{center}
\end{figure}

@@

By \textit{linked data sets} we mean that the data have more than one
conceptual grouping and that there is a relationship between these
different conceptual groupings.  Linked data sets are 
familiar to users of relational databases.  Data in a relational
database are typically stored in many linked tables with rows in
different tables linked to each other through a unique identifier
called a key.  In this example, each table can be considered a data set
and they are linked through the keys. 
Linked relational database tables could be combined into one
large table, but doing so would result in a large amount of redundancy if
the elements in the different tables are linked in one-to-many or many-to-many
relationships.  Instead in relational databases, the tables are created such
that each data element only needs to be stored in one table and then is linked
to data elements in other tables.  Our design will build on this
concept, but the idea is more general and we shall discuss other situations
where the linked data sets do not easily correspond to tables.  As an example
suppose we have gene expression data and we
also know the Gene Ontology (GO) molecular function terms for these genes. 
These data have a many-to-many relationship where each gene can be
annotated at more than one GO term and each GO term can describe more than one
gene.  Now these data could be combined in one data structure, but in
our design we find that creating a MVC for the gene expression data and a MVC
for the GO graph allows us to create views where we can explore the
relationships between the genes and the GO terms.  To explore these
relationships between data elements in different MVCs, we
must have a map that goes from the entities in one data set to those in the
other, as well as the inverse map, in order to propagate signals between the
MVCs.

Consider Figure \ref{Fig:firstMP}, which provides a cartoon
description of the steps that occur when a user interacts with a
view.  In the cartoon there are only two views, one of model A (note 
that once a data set is loaded in a MVC, we call it a model) and the
other of the linked model B.  When a user
interacts with the scatter plot view of model B, the response to
this interaction is for model B to update itself (this is indicated
on the figure as a circled 1).  In Step 2, the scatter plot view is
told to update itself.  Next, in Step 3, because model B is linked
to model A, model B sends a message to model A.  In Step 4,
the information from model B is converted into information that
model A understands.  In Step 5, this information is received by
model A and model A updates itself.  Finally, in Step 6, model
A tells its view to update itself.

@@

%%FIXME: please update the figure so that it has A and B, not 1 and 2
%%for the data sets - it was too confusing
%%EAW: it's changed
\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=3.3in, width=6in]{firstMessagePassing.jpg}
    \caption{ Linked views of linked models.  When a user interacts with
      the y versus x scatter plot, the first step, indicated by a circled one,
      is to tell model B to update.  Then all views of model B must be
      updated.  Next any data that are linked to model B must be notified
      that model B changed.  This information from model B must be
      altered so that it can be understood by linked models, which is shown
      in the fourth step.  Finally, model A is notified that it should
      update in response to the change in model B and then all views of
      model A are updated. }
    \label{Fig:firstMP}
  \end{center}
\end{figure}

@@

In essence the MVC design assumes that there is one data set, which is
called the model, that all views reference.  But problems arise when
the data consist of more than one conceptual grouping.  For example,
consider data stored in a relational database, such as the data shown in
Figure \ref{FigDBTab}.  Here the database
consists of three tables, State, City and StateCity.  The State table
has three columns, State (the key), Life Expectancy, and Income.  The
City table has three columns, City (the key), Precipitation, and
Temperature.  The StateCity table has two columns, State and City.
This last table relates which cities are in which states.  In this
example, each city can be in only one state, but each state can have
more than one city so there is a one-to-many relationship between
states and cities.  In order to support interactive views of all data
tables, in a standard MVC paradigm, it seems that either all data must
be combined into a single large table (Figure \ref{Fig:OneDBTab}), or
the processing and accessing of the model data must become much more
complex.

%where there are
%multiple tables of data and rows in different tables are linked
%through a unique identifier called a key.  Although these tables of
%data could be combined into one large table, there are several reasons
%not to do this.  Keeping the data in separate tables reduces
%redundancy, particularly when the rows are related by a one-to-many or
%a many-to-many relationship.

@@

%Consider Figure \ref{Fig:DBTab} shown below.  Here the database
%consists of three tables, State, City and StateCity.  The State table
%has three columns, State (the key), Life Expectancy, and Income.  The
%City table has three columns, City (the key), Precipitation, and
%Temperature.  The StateCity table has two columns, State and City.
%This last table relates which cities are in which states.  In this
%example, each city can be in only one state, but each state can have
%more than one city so there is a one-to-many relationship between
%states and cities.  In order to support interactive views of all data
%tables, in a standard MVC paradigm, it seems that either all data must
%be combined into a single large table (Figure \ref{Fig:OneDBTab}), or
%the processing and accessing of the model data must become much more
%complex.

%\begin{figure}[ht]
%  \begin{center}
%    \includegraphics[height=0.9in, width=3.5in]{newDatabaseTables.jpg}
%    \caption{ Example of relational database tables. }
%    \label{Fig:DBTab}
%  \end{center}
%\end{figure}

%\begin{figure}[ht]
%  \begin{center}
%    \includegraphics[height=1.2in, width=1.5in]{newOneDatabaseTable.jpg}
%    \caption{ Example of a combined table. }
%    \label{Fig:OneDBTab}
%  \end{center}
%\end{figure}

\begin{figure}[ht]
  \begin{center}
    \subfigure[Example of relational database
    tables.]{
    \label{Fig:DBTab}
    \includegraphics[height=1.2in, width=3.5in]{newDatabaseTables.jpg}}
    \hspace{.3in}
    \subfigure[Example of a combined table.]{
    \label{Fig:OneDBTab}
    \includegraphics[height=1.2in, width=1.5in]{newOneDatabaseTable.jpg}}
    \caption{Relational database tables.}
%    \label{Fig:DBTab}
  \end{center}
\end{figure}

We felt that neither approach had much appeal.  A single large table
suggests that all data are similar, and still leaves issues about how
to deal with ensuring that when, for example, a state is selected all
cities within it are highlighted.  The second approach requires writing a
large amount of code, but is unlikely to be extensible.
We should also mention that our main area of
application is a setting where one representation of the data is as a
graph (with nodes and edges) while the other is as a matrix. We can
envisage lots of different MVCs that operate on graph based data, and
equally MVCs that operate on matrix based data. Specifying a mechanism
by which these can interact allows us to create two visible classes
that others can use, but still provide interaction between
appropriately linked instances of the two classes. We also argue that
there are surely other specific types of models that represent
specific types of data (e.g. time series) and it would make sense to
implement these as yet another separate, specialized MVC, that could
be linked to any one of a variety of other MVCs.

@@

As discussed in Section \ref{Sec:Intro}, the MVC design typically uses a single
conceptual unit, such as a data frame or a matrix, for the model.  And
we believe that such an approach is a good one.  MVCs should be
specialized to certain types of data, as this allows them to provide
better and more natural interactions.  However, we additionally argue
that there is great benefit to considering linking these different
MVCs when the underlying data themselves are derived from the same source.
We believe that there will often be a hierarchical relationship
between the models and the MVCs that contain them.  We have found
that allowing the user to create new MVCs based on a current MVC is an
important and useful paradigm.

Each \Robject{MVC} is a coherent object and the interactions between
\Robject{MVCs} are encapsulated in a set of messages.  We use one
\Robject{MVC} object to handle each model.  Thus, within a
\Robject{MVC} object, one can create linked views of that particular
model, using the standard MVC design.  Between \Robject{MVC}
objects we propose a second messaging system that allows messages to
be sent between linked models.

@@

In Figure \ref{Fig:MultMVC} we show one view for state-level data and
one view for the city-level data.  The state and city data are linked
based on the color.  Even though there are two \Robject{MVC} objects
here, one \Robject{MVC} for each model, the views are linked.

@@

@@

In the MVC design, the model stores the data and among other things it
supports all updating and other mechanisms that might be needed.
Information can be sent to the model to add, update, and remove data
and after every change the model is responsible for sending messages
to the views indicating which changes have occurred.  In our
implementation we created a single virtual class that represents an
abstract model and that has storage locations for data that are
needed by all models.  Then, we created a number of subclasses, one for
each type of model.  Currently, we have implemented three types of models, 
one based on data frames, one based on graphs, and one based on the 
expression set class of data used by the Bioconductor project \citep{BIOC}.

%In Figure~\ref{Fig:Model} we show the inheritance
%structure for the three types of models we have implemented, one based
%on data frames, one based on graphs and one based on the expression
%set class of data used by the Bioconductor project \citep{BIOC}.

%%FIXME: we need to send the complete paper2 to the BioC tech report
%%series and then cite it here, saying something like: Complete
%%implementation details are provided in XXX.

@@

%\begin{figure}[ht]
%  \begin{center}
%    \includegraphics[height=3in, width=4.7in]{ModelClass.jpg}
%    \caption{ Inheritance for Model Objects. }
%    \label{Fig:Model}
%  \end{center}
%\end{figure}

@@

The view classes represent the visual depictions, or renderings, of
the model.  There are some data that are required by all views, and all views
must respond to a certain subset of events so we have implemented the view
classes by first creating a single, virtual class which the specialized
views extend.  As with our model classes, this allows us to define
methods and data storage once, for the virtual class, and only extend
it for views that specifically require additional data storage, or
specialized handling.  The object model for the view classes is shown
in Figure \ref{Fig:View}.

@@

The controller performs two tasks: the first task is to maintain state,
such as what response will occur for a particular user interaction, 
and the second task is to control message passing.  Messages, which are
discussed in Sections \ref{Ssec:OneMess} and \ref{Ssec:MultLink}, are
responsible for creating new components, such as models and views, and for
passing information between components, such as a view asking a model for
information.  The controller determines when these messages are created and
ensures that they are sent to the proper recipient.

For the second task, the controller gives the user restricted access
to the state information through an application program interface
(API) and thereby lets the user change the response to different events. 
In our implementation users can interact either through a GUI
or through command line functions.  Through the API the user can load
data, create views, set the response to user interaction with a view,
quit the program and perform other functions.  On the GUI, which we call the
control window, users can decide which action is taken next by interacting
with one of the menus: \texttt{File}, \texttt{Edit},
\texttt{Display}, \texttt{Event}, and \texttt{Window}.  If the user
would rather not use the GUI, there are command line functions that
provide access to the same functionality that is available through the
GUI.

@@

We support multiple loaded \Robject{MVCs}, but the control window menus
and the command line functions only perform operations on one
\Robject{MVC} object at a time.  Thus, we make use of the concept of
an \textit{active} \Robject{MVC} object, where the active
\Robject{MVC} object (and thus, the active model) means that
operations performed through the control window menus or through the
command line functions affect that \Robject{MVC} object. Users can
switch between active models, but since the interactions could be
quite different, there must always be an active MVC.

Although the same menus always appear on the control window,
the items under each menu may change depending on the type of model
that is active.  For example, the \texttt{Display} menu allows users
to create views of the active model, but only certain views are
appropriate for a type of model, as mentioned in
Section~\ref{Ssec:OneViews}.  If the active model was a data frame
model (\Rclass{dfModel}), then the views that the user could create
are a scatter plot (\Rclass{sPlotView}) and a spreadsheet
(\Rclass{spreadView}), but a graph plot (\Rclass{graphView}) does not
make sense.  Thus, when the model in the \textit{active} \Robject{MVC} is
of class \Rclass{dfModel}, then the Display menu has menu items to
create a scatter plot and a spreadsheet.

Because one of our goals is to create a software package that is
flexible and extensible, users and developers can add menus and menu items 
to the control window through the command line to access new
functionality that they have created.  As an example if a future
programmer wants to add a new type of view, such as a histogram, then
a new menu item for that view could be added to the \texttt{Display}
menu.  When adding to the API, command line functions should be created as
well as menu items to access the new functionality.

@@

Linking callback functions to events, such as a button click, a mouse
movement or a key press event, is what allows users to have an
interactive environment.  When an event occurs, the callback function
linked to the event determines the response to that event.  To create
a flexible environment it is important to be able to change the
response to an event.  For example, when a left button press event
occurs, a user may sometimes want a point to be colored and at other
times she may want a point to be hidden.  However, different responses
to an event may require functions with different signatures, where a function
signature is the combination of the function name along with its parameters.
In our example coloring a point requires two parameters, the new color to use
and the point to color, while hiding a point requires just one parameter, the
point to hide.

@@

%%FIXME: I have tried to reword this part, but am not sure I have it
%%right, nor that I am following the design decision here.
%% EAW: When we decided that we wanted callback functions to be
%% interchangeable, we decided that the callback functions must have the same
%% signature so that calls to different callback functions would be
%% identical, as we wouldn't know ahead of time what function we would be
%% calling.  All callback functions must know which object in the plot to
%% act on, but there was no other universal information that all callback
%% functions needed.  Thus, for any other information that a callback function
%% would need, preprocessing functions were used.  As for that one parameter
%% that the callback function needed, the parameter type may be different
%% depending on the plot (and model) type - whatever is considered the object
%% of interest in a plot (and model).  Thus, for plots of data frames,
%% the object in the plot is considered a row from the data frame and for
%% plots of graphs, the object in the plot is considered a node in the graph. 
Since the intention is to be able to interchange the responses to an event,
the signatures of different callback functions should be the same.
In our design a callback function takes only one parameter. 
The class of that parameter depends on the type of model represented by the
\Robject{MVC}.  For example, for a graph model, the parameter corresponds to
the name of a node in the graph, while for a data frame model, the parameter
represents a case.  We expect a one-to-one relationship between the elements
on a view and the elements in the model.  Continuing the previous example,
each point on a scatter plot represents one case in the data frame model and
each node in the graph plot represents one node in the graph model.  Clearly
it is possible for two points in a scatter plot to overlap if they have the
same x axis and y axis values, but they are still considered two points
representing two cases in the model.

%%FIXME: I thought we talked about this, and found it to be a little
%%problematic, as the way this sounds every click will ask me to set a
%%color. Somehow, some of these are essentially global options that
%%need to be set once, and then used.
%% EAW: The preprocessing functions are called once when the callback function
%% is linked to the event - they are not called each time the event occurs.
Any other information a callback function may need is set using the
preprocessing functions, which are automatically called once by the controller
when the callback
function is linked to the event.  The preprocessing functions take no
parameters.  The purpose of the preprocessing functions is to set
variables that the callback function needs.  Thus, not all callback
functions need preprocessing functions.  If the callback
function sets the color of a node in a graph, then a preprocessing
function is needed to determine what color is used.  So in this
example, the preprocessing function opens a color browser to allow the
user to choose the new color.

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=4in, width=4.5in]{MVCMessageClass2.jpg}
    \caption{ Inheritance for Message Classes that are Passed Between
      \Robject{MVCs}. }
    \label{Fig:BetwMess}
  \end{center}
\end{figure}
